Exploring the Pareto front of multi-objective COVID-19 mitigation policies using reinforcement learning

### 2023-02-09 paper/main.tex
remove \added

\newcommand{\foi}{\lambda}
\newcommand\added[1]{\textcolor{blue}{#1}}
\newcommand\todo[1]{\textcolor{red}{TODO: #1}}

@@

\added{In a classical MDP, executing an action $\action_t$ in any state $\state_t$ leads to a next state $\state_{t+1}$ according to the transition function $\mdptransition$. At every timestep $t$, the policy is free to choose the action to perform. In our case, this potentially results in a different restriction $[p_{w}, p_{s}, p_{l}]$ every week. However, we argue that in the context of mitigation policies, consistency is key, and policies that impose changes too frequently will be hard to follow.}

\added{In order to learn realistic and consistent mitigation policies, we incorporate a \emph{budget} on the number of times a policy can change its actions until the terminal state of the MOMDP. Concretely, when the action changes, i.e., if the social restriction proposed by the policy is different from the current one in place, we reduce the budget by one. We only allow action changes as long as there is budget left. We note that we can simulate a no-limit budget setting by setting the budget to the length of the episode.}

@@

\paragraph{State-space:} The state of the MOMDP \added{concerns a 3-tuple. The first element,} $\mdpstate_m$, directly corresponds to the aggregation of the state variables in the epidemiological model, i.e., a tuple,
\begin{equation}
    \{S_k, E_k, I_k^{presym}, I_k^{asym}, I_k^{mild}, I_k^{sev}, I_k^{hosp}, I_k^{icu}, H^{new}_k, D_k, R_k\},
\end{equation}
%
for each age group $k \in \{1, \ldots, \agegroups\}$, where $S$ encodes the members of the population who are susceptible to infection and $E$ encodes the members of the population who have been exposed to COVID-19. Moreover, $I^{presym}$, $I^{asym}$, $I^{mild}$, $I^{hosp}$, $I^{icu}$ are the members of the population infected with COVID-19 and are, respectively, presymptomatic, asymptomatic, have mild symptoms, are hospitalised, or are in the ICU. Finally, $H^{new}_k$ represents the number of newly hospitalised individuals in age group $k$.

We parameterize the \added{epidemiological} model using the mean of the posteriors as specified by \citet{abrams2021modelling} (details in SI).

\added{The second element of the tuple consists of the social contact matrix $\hat{C}$ that is currently in place. The reason to incorporate it in the state-space is two-fold. First, \citet{abrams2021modelling} define a compliance function, simulating the time people need to get used to the new rules set in place. As such, during the simulated week, there is a gradual shift from the current $\hat{C}$ to the next one. We thus require to know the current $\hat{C}$.}

\added{Secondly, we require the current $\hat{C}$ to determine whether the action changes the social restrictions in place -- and thus consume part of the budget -- or not.}

\added{The third element of the tuple concerns of the budget $\budget$. We note that we incorporate a distinct budget per action-dimension, so $p_{w}, p_{s}$ and $p_{l}$ each have their own budget, rendering $\budget$ a vector $\budget = [b_w, b_s, b_l]$. As such, it is possible that, at timestep $t$, the budget for one of the proportional reductions is reduced but not the others.}

@@

\added{TODO REPHRASE, FROM PCN PAPER: Using neural networks as function approximators in RL comes with many challenges. One of them is that the target (e.g., the optimal action of the policy) is not known in advance --- as opposed to classical supervised learning where the ground-truth target is provided. As the behavior of the agent improves over time, the action used as target can change, often leading to hard-to-tune and brittle learners \cite{mnih2015,fu2019}.}

\added{Instead of trying to continuously improve the policy by learning actions that should lead to the highest cumulative reward, PCN flips the problem, by learning actions that should lead to any \emph{desired} cumulative reward (be it high or low). In this way, all past trajectories can be reused for supervision, since their returns are known, as well as the actions needed to reach said returns. We can thus train a policy that, conditioned on a desired return, provides the optimal action to reach said return. By leveraging the generalization properties of neural networks, we can accumulate incrementally better experience by conditioning on increasingly higher reward-goals}

@@

Similar to classification, PCN requires a labeled dataset with training examples to learn a mapping from input to label. However, contrary to classification, the data present in the dataset is not fixed. \added{PCN collects data from the trajectories experienced while exploring the environment. Thus, the dataset improves over time, as we collect increasingly better trajectories.}

\added{Concretely, the PCN algorithm is a loop that repeatedly executes 3 main steps:
\begin{enumerate}
    \item Use the current policy to interact with the environment and generate trajectories (see Section~\ref{sec:pcn-policy-exploration}).
    \item Update the dataset to incorporate these new trajectories and prun the least interesting trajectories in terms of returns (see Section~\ref{sec:pcn-building-dataset}).
    \item Update the policy by training it on the updated dataset (see Section~\ref{sec:pcn-continuous}).
\end{enumerate}
}

@@

\added{Given a trajectory composed of $T$ transitions $(s_0, a_0, \bm{r}_0, s_1), \dots, (s_{T-1}, a_{T-1}, \bm{r}_{T-1}, s_T)$ we can compute, for each transition at timestep $t$, the future discounted return $\bm{R}_t = \sum_{i=t}^T \gamma^i\bm{r}_i$ and the leftover horizon $h_t = T-t$. Since for this trajectory executing action $a_t$ in state $s_t$ resulted in return $\mathbf{R}_t$ in $h_t$ timesteps, we add a datapoint with input $\langle s, \hat{h}, \mathbf{\hat{R}}\rangle = \langle s_t, h_t, \mathbf{R}_t \rangle$ and output $a=a_t$ to the dataset. In other words, when the observed return corresponds to the desired return in that state, then $a_t$ is the optimal action to take.}

\added{Since we have a fix-sized dataset, adding this trajectory means we need to remove another one. Our aim is to keep trajectories that span different parts of the objective space, such that we can explore as many types of trade-offs as possible. However, to improve the trade-offs we currently have in our dataset, we need to only keep solutions that are close to our current estimate of the Pareto front. Thus, we assign a priority for each trajectory in our dataset that combine two metrics: the \emph{crowding distance}, which measures the distance of a solution with its closest neighbors, and the \emph{l2-norm} with the closest non-dominated solution, which indicates how close the solution is from our current estimate of the Pareto front.}

\added{The dataset is then updated by only keeping the trajectories with the top-$N$ priorities.}

@@

\added{In the previous section, we explain how PCN updates the dataset given a trajectory. In this section, we explain how PCN produces said trajectory.}

\added{It is unrealistic to expect PCN to reliably produce trajectories with high-valued desired returns when it has only been trained on datapoints originating from random trajectories. Rather, we can expect PCN to produce trajectories with returns in the range of the ones from the current training data. Therefore, if we obtain trajectories reaching high returns, PCN will be able to confidently return high-return policies.}

\added{PCN leverages the fact that, due to the generalization capabilities of neural networks, the policies obtained from the network will still be reliable even if the desired return is marginally higher than what is present in the training data. In fact, they will perform similar actions to those in the training data, but lead to a higher return. Thus, we incrementally condition the network on better and better returns, in order to obtain trajectories that extend the boundaries of PCN's current coverage set.}

\added{More precisely, we randomly select a non-dominated return $\mathbf{R}_{nd}$ and its corresponding horizon $\hat{h}$ from the dataset. By randomly picking a non-dominated return from the entire coverage set we ensure equal chance of improvement to each part of the objective space. To generate trajectories that improve upon $\mathbf{R}_{nd}$, we randomly select an objective $o$ and increase $R_{nd,o}$ by a sample from the uniform distribution $U(0, \sigma_o)$, where $\sigma_o$ is the standard deviation for the selected objective, using all non-dominated returns from the trajectories in the dataset. This then becomes our desired return $\bm{\hat{R}}$. By restricting the improvement to at most $\sigma_o$, $\mathbf{\hat{R}}$ stays in the range of possible achievable returns and, by only modifying one objective at a time, the changes to the network's input compared to the training data are kept at a minimum.}

@@

Even with the $\mdprewardfn_\text{ARI}$, the increased budget shows an increase in hypervolume in terms of hospitalisations. Moreover, those hypervolumes are close to the ones trained on $\mdprewardfn_\text{ARH}$, inidicating that their coverage sets are similar. However, regardless of the imposed budget, the hypervolumes are slightly worse. This is to be expected, since those experiments are not directly optimising on $\mdprewardfn_\text{ARH}$. We make similar conclusions for $I_\varepsilon$. For budgets 2, 3 and 5, the difference between the worst-performing policy for the $\mdprewardfn_\text{ARI}$ variant and the $\mdprewardfn_\text{ARH}$ is less than $0.01$, indicating less than $1\%$ difference in return values between the two variants when comparing their worst-performing policy. As an exception, we notice that PCN without budget restrictions results in better performance across every metric for the $\mdprewardfn_\text{ARI}$ variant. Still, due to the high standard deviation of the no-budget, $\mdprewardfn_\text{ARH}$ setting, we do not believe this difference is meaningfull. \todo{some kind of meaningfull conclusion}

>>>>

\newcommand{\foi}{\lambda}
\newcommand\todo[1]{\textcolor{red}{TODO: #1}}

@@

In a classical MDP, executing an action $\action_t$ in any state $\state_t$ leads to a next state $\state_{t+1}$ according to the transition function $\mdptransition$. At every timestep $t$, the policy is free to choose the action to perform. In our case, this potentially results in a different restriction $[p_{w}, p_{s}, p_{l}]$ every week. However, we argue that in the context of mitigation policies, consistency is key, and policies that impose changes too frequently will be hard to follow.

In order to learn realistic and consistent mitigation policies, we incorporate a \emph{budget} on the number of times a policy can change its actions until the terminal state of the MOMDP. Concretely, when the action changes, i.e., if the social restriction proposed by the policy is different from the current one in place, we reduce the budget by one. We only allow action changes as long as there is budget left. We note that we can simulate a no-limit budget setting by setting the budget to the length of the episode.

@@

\paragraph{State-space:} The state of the MOMDP concerns a 3-tuple. The first element, $\mdpstate_m$, directly corresponds to the aggregation of the state variables in the epidemiological model, i.e., a tuple,
\begin{equation}
    \{S_k, E_k, I_k^{presym}, I_k^{asym}, I_k^{mild}, I_k^{sev}, I_k^{hosp}, I_k^{icu}, H^{new}_k, D_k, R_k\},
\end{equation}
%
for each age group $k \in \{1, \ldots, \agegroups\}$, where $S$ encodes the members of the population who are susceptible to infection and $E$ encodes the members of the population who have been exposed to COVID-19. Moreover, $I^{presym}$, $I^{asym}$, $I^{mild}$, $I^{hosp}$, $I^{icu}$ are the members of the population infected with COVID-19 and are, respectively, presymptomatic, asymptomatic, have mild symptoms, are hospitalised, or are in the ICU. Finally, $H^{new}_k$ represents the number of newly hospitalised individuals in age group $k$.

We parameterize the epidemiological model using the mean of the posteriors as specified by \citet{abrams2021modelling} (details in SI).

The second element of the tuple consists of the social contact matrix $\hat{C}$ that is currently in place. The reason to incorporate it in the state-space is two-fold. First, \citet{abrams2021modelling} define a compliance function, simulating the time people need to get used to the new rules set in place. As such, during the simulated week, there is a gradual shift from the current $\hat{C}$ to the next one. We thus require to know the current $\hat{C}$.

Secondly, we require the current $\hat{C}$ to determine whether the action changes the social restrictions in place -- and thus consume part of the budget -- or not.

The third element of the tuple concerns of the budget $\budget$. We note that we incorporate a distinct budget per action-dimension, so $p_{w}, p_{s}$ and $p_{l}$ each have their own budget, rendering $\budget$ a vector $\budget = [b_w, b_s, b_l]$. As such, it is possible that, at timestep $t$, the budget for one of the proportional reductions is reduced but not the others.

@@

\todo{TODO REPHRASE, FROM PCN PAPER} Using neural networks as function approximators in RL comes with many challenges. One of them is that the target (e.g., the optimal action of the policy) is not known in advance --- as opposed to classical supervised learning where the ground-truth target is provided. As the behavior of the agent improves over time, the action used as target can change, often leading to hard-to-tune and brittle learners \cite{mnih2015,fu2019}.

Instead of trying to continuously improve the policy by learning actions that should lead to the highest cumulative reward, PCN flips the problem, by learning actions that should lead to any \emph{desired} cumulative reward (be it high or low). In this way, all past trajectories can be reused for supervision, since their returns are known, as well as the actions needed to reach said returns. We can thus train a policy that, conditioned on a desired return, provides the optimal action to reach said return. By leveraging the generalization properties of neural networks, we can accumulate incrementally better experience by conditioning on increasingly higher reward-goals

@@

Similar to classification, PCN requires a labeled dataset with training examples to learn a mapping from input to label. However, contrary to classification, the data present in the dataset is not fixed. PCN collects data from the trajectories experienced while exploring the environment. Thus, the dataset improves over time, as we collect increasingly better trajectories.

Concretely, the PCN algorithm is a loop that repeatedly executes 3 main steps:
\begin{enumerate}
    \item Use the current policy to interact with the environment and generate trajectories (see Section~\ref{sec:pcn-policy-exploration}).
    \item Update the dataset to incorporate these new trajectories and prun the least interesting trajectories in terms of returns (see Section~\ref{sec:pcn-building-dataset}).
    \item Update the policy by training it on the updated dataset (see Section~\ref{sec:pcn-continuous}).
\end{enumerate}

@@

Given a trajectory composed of $T$ transitions $(s_0, a_0, \bm{r}_0, s_1), \dots, (s_{T-1}, a_{T-1}, \bm{r}_{T-1}, s_T)$ we can compute, for each transition at timestep $t$, the future discounted return $\bm{R}_t = \sum_{i=t}^T \gamma^i\bm{r}_i$ and the leftover horizon $h_t = T-t$. Since for this trajectory executing action $a_t$ in state $s_t$ resulted in return $\mathbf{R}_t$ in $h_t$ timesteps, we add a datapoint with input $\langle s, \hat{h}, \mathbf{\hat{R}}\rangle = \langle s_t, h_t, \mathbf{R}_t \rangle$ and output $a=a_t$ to the dataset. In other words, when the observed return corresponds to the desired return in that state, then $a_t$ is the optimal action to take.

Since we have a fix-sized dataset, adding this trajectory means we need to remove another one. Our aim is to keep trajectories that span different parts of the objective space, such that we can explore as many types of trade-offs as possible. However, to improve the trade-offs we currently have in our dataset, we need to only keep solutions that are close to our current estimate of the Pareto front. Thus, we assign a priority for each trajectory in our dataset that combine two metrics: the \emph{crowding distance}, which measures the distance of a solution with its closest neighbors, and the \emph{l2-norm} with the closest non-dominated solution, which indicates how close the solution is from our current estimate of the Pareto front.

The dataset is then updated by only keeping the trajectories with the top-$N$ priorities.

@@

In the previous section, we explain how PCN updates the dataset given a trajectory. In this section, we explain how PCN produces said trajectory.

It is unrealistic to expect PCN to reliably produce trajectories with high-valued desired returns when it has only been trained on datapoints originating from random trajectories. Rather, we can expect PCN to produce trajectories with returns in the range of the ones from the current training data. Therefore, if we obtain trajectories reaching high returns, PCN will be able to confidently return high-return policies.

PCN leverages the fact that, due to the generalization capabilities of neural networks, the policies obtained from the network will still be reliable even if the desired return is marginally higher than what is present in the training data. In fact, they will perform similar actions to those in the training data, but lead to a higher return. Thus, we incrementally condition the network on better and better returns, in order to obtain trajectories that extend the boundaries of PCN's current coverage set.

More precisely, we randomly select a non-dominated return $\mathbf{R}_{nd}$ and its corresponding horizon $\hat{h}$ from the dataset. By randomly picking a non-dominated return from the entire coverage set we ensure equal chance of improvement to each part of the objective space. To generate trajectories that improve upon $\mathbf{R}_{nd}$, we randomly select an objective $o$ and increase $R_{nd,o}$ by a sample from the uniform distribution $U(0, \sigma_o)$, where $\sigma_o$ is the standard deviation for the selected objective, using all non-dominated returns from the trajectories in the dataset. This then becomes our desired return $\bm{\hat{R}}$. By restricting the improvement to at most $\sigma_o$, $\mathbf{\hat{R}}$ stays in the range of possible achievable returns and, by only modifying one objective at a time, the changes to the network's input compared to the training data are kept at a minimum.

@@

Even with the $\mdprewardfn_\text{ARI}$, the increased budget shows an increase in hypervolume in terms of hospitalisations. Moreover, those hypervolumes are close to the ones trained on $\mdprewardfn_\text{ARH}$, inidicating that their coverage sets are similar. However, regardless of the imposed budget, the hypervolumes are slightly worse. This is to be expected, since those experiments are not directly optimising on $\mdprewardfn_\text{ARH}$. We make similar conclusions for $I_\varepsilon$. For budgets 2, 3 and 5, the difference between the worst-performing policy for the $\mdprewardfn_\text{ARI}$ variant and the $\mdprewardfn_\text{ARH}$ is less than $0.01$, indicating less than $1\%$ difference in return values between the two variants when comparing their worst-performing policy. As an exception, we notice that PCN without budget restrictions results in better performance across every metric for the $\mdprewardfn_\text{ARI}$ variant. Still, due to the high standard deviation of the no-budget, $\mdprewardfn_\text{ARH}$ setting, we do not believe this difference is meaningfull. \todo{some kind of meaningful conclusion}
